---
title: "Principal Components Analysis"
output: html_document
---

This document describes the process of running PCA for *Geleznowia* ingroup samples using strict-filtered SNPs

First, set the working directory to the folder with the input files (e.g. a link called `input.vcf` for the `set1.vcf` file)

Load up the libraries and designate the input files  
```{r}
suppressMessages(library(adegenet))
suppressMessages(library(vcfR))
vcf_file <- "input.vcf"
```

Create colour palettes and point choices  
These should be as long as the number of main groups to designate  
```{r}
col_1 <- "#218a21"
col_2 <- "#b8a729"
col_3 <- "#ff8c00"
col_4 <- "#2e7eff"
col_5 <- "#c637ff"
col_pal <- c(col_1, col_2, col_3, col_4, col_5)

# set a second colour to create a range with the first
col_1b <- "#2cff2c"
col_2b <- "#ffe946"
col_3b <- "#bd2e0e"
col_4b <- "#15438e"
col_5b <- "#540573"
col_palb <- c(col_1b, col_2b, col_3b, col_4b, col_5b)

# set point choices for the groups
# 21 - filled circle, 22 - filled square, 23 - filled diamond,
# 24 - filled triangle, 25 - filled triangle (down)
point_ch <- c(21, 22, 23, 24, 25)
```

Create a group sampling scheme for morphological identitifications  
These will be searched against sample names  
```{r}
ama <- c("AMA")
bin <- c("BIN")
mar <- c("MAR02", "MAR03", "MAR04", "MAR05", "MAR08", "MAR10")
ver <- c("VER", "MAR01", "MAR06", "MAR07", "MAR09")
wht <- c("WHT")
groups <- list(ama, bin, mar, ver, wht)
```

Create a function for constructing a dataframe to hold groups, colours and points  
```{r}
make_df <- function(samples, groups, col_pal1, col_pal2, point_ch) {

	# define a function to generate colours between two extremes
	get_colours <- function(col1, col2, numsteps) {
		rangefunc <- colorRampPalette(c(col1, col2))
		colorrange <- rangefunc(numsteps)
		colorrange[seq_len(numsteps)]
	}

	# set up the dataframe
	mydf <- as.data.frame(matrix(ncol = 4, nrow = length(samples)))
	colnames(mydf) <- c("sample", "group", "colour", "point")
	mydf$sample <- samples

	# cycle through the groups
	indices <- rep(0, length(samples))
	index <- 1
	for (group in groups) {
		search_term <- paste(group, collapse = "|")
		indices[grep(search_term, samples)] <- index
		sample_list <- samples[grep(search_term, samples)]

		# determine how many colours are needed (unique pops in the group)
		# pops are named by the first part of the sample name before "-"
		x <- sapply(sample_list, function(x) strsplit(x, "-")[[1]][1], USE.NAMES = FALSE)
		pops <- unique(x)

		# generate the colours
		if (length(pops) > 1) {
			these_colours <- get_colours(col_pal1[index], col_pal2[index], length(pops))
		} else {
			these_colours <- list(col_pal1[index])
		}

		# assign them
		index2 <- 1
		for (pop in pops) {
			mydf$colour[grep(pop, samples)] <- these_colours[[index2]]
			index2 <- index2 + 1
		}

		# increment the group index
		index <- index + 1
	}

	# assign group number, point and colours
	mydf$group <- indices
	mydf$point <- point_ch[indices]

	# return the dataframe
	mydf
}
```

Create a function to run and plot a PCA
```{r}
run_pca <- function(genl, mydf, x = 1, y = 2) {
	# run the PCA and grab elements
	pca <- glPca(genl, nf = 5, parallel = TRUE)
	pca_var <- round(100 * pca$eig / sum(pca$eig), digits = 1)
	method <- paste0("glPca (", nLoc(genl), " SNPs)")
	pca_scores <- pca$scores

	# grab plotting features from the dataframe
	colours <- mydf$colour
	point_type <- mydf$point

	# plot
	par(mar = c(5.1, 6.1, 5.1, 2.1), cex.lab = 2)
	plot(pca_scores[, x], pca_scores[, y],
		xlab = paste0("PC", x, " (", pca_var[x], "%)"),
		ylab = paste0("PC", y, " (", pca_var[y], "%)"),
		main = paste0("Principal Component Analysis\n", method),
		bg = colours,
		col = "black",
		lwd = 0.25,
		pch = point_type,
		asp = 1,
		cex = 2,
		panel.first = {
			grid()
			abline(h = 0, v = 0)
		},
		cex.axis = 1.5,
		cex.main = 1.5)
	par(mar = c(5.1, 4.1, 4.1, 2.1), cex.lab = 1)
}
```

Create a function to plot the legend
```{r}
plot_legend <- function(mydf, myorder) {
	leg_df <- unique(mydf[, 2:4])
	leg_df <- leg_df[match(myorder, leg_df$group), ]
	plot(1, type = "n", axes = FALSE, xlab = "", ylab = "")
	legend("center", legend = leg_df$group,
		pch = leg_df$point, pt.bg = leg_df$colour,
		col = "black", title = NULL, cex = 1.5,
		pt.cex = 2, pt.lwd = 0.25)
}
```

Read in the VCF  
```{r}
vcf <- read.vcfR(vcf_file, verbose = FALSE)
cat("Read in a VCF with", ncol(vcf@gt) - 1, "samples,",
	length(unique(vcf@fix[, 1])), "loci and", nrow(vcf@fix), "SNPs\n")
genl <-	vcfR2genlight(vcf)
```

Generate the dataframe with assigned colours and points for each sample  
```{r}
samples <- indNames(genl)
mydf <- make_df(samples, groups, col_pal, col_palb, point_ch)
```

Start creating a single pdf
```{r}
pdf("pca.pdf", width = 7, height = 7)
```

Run and plot the PCA
```{r}
run_pca(genl, mydf, x = 1, y = 2)
run_pca(genl, mydf, x = 1, y = 3)
run_pca(genl, mydf, x = 1, y = 4)
run_pca(genl, mydf, x = 2, y = 3)
run_pca(genl, mydf, x = 2, y = 4)
run_pca(genl, mydf, x = 3, y = 4)
plot_legend(mydf, c(1, 2, 3, 5, 4))
```

Stop creating the pdf
```{r}
invisible(dev.off())
```

If interested, the PCA can be plotted interactively
```{r}
suppressMessages(library("plotly"))
pca <- glPca(genl, nf = 5, parallel = TRUE)
data <- as.data.frame(pca$scores)
x_vals <- data$PC1
y_vals <- data$PC2
z_vals <- data$PC3
text_field <- rownames(data)
```

```{r}
plot_ly(data = data, type = "scatter", mode = "markers",
		x = ~x_vals, y = ~y_vals, text = text_field, size = 15
)
```

```{r}
plot_ly(data = data, type = "scatter3d", mode = "markers",
		x = ~x_vals, y = ~y_vals, z = ~z_vals,
		text = text_field, size = 15
)
```
